---
title: "Data Vizualization 2 - Homework 1"
author: "Nawal Zehra Hasan"
date: "1/16/2022"
output:
  html_document:
    df_print: paged
  rmarkdown::html_document:
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q0. Load the nycflights13 package and check what kind of datasets exist in the package, then create a copy of flights dataset into a data.table object, called flight_data.

```{r, message=FALSE, warning=FALSE,include=FALSE}
#cleaning the environment
rm(list=ls())
#loading relevant packages
library(data.table)
library(nycflights13)
library(lubridate)
library(ggplot2)
library(cowplot)
library(rgeos)
library(rnaturalearthdata)
library(googleway)
library(ggrepel)

#load nycflights data
flight_data <- data.table(nycflights13::flights)
#check the data type
class(flight_data) 
#check the structure
str(flight_data)
```

```{r, include=TRUE, message=TRUE, echo=FALSE}
head(flight_data)
```


## Q1. Which destination had the lowest avg arrival delay from LGA with minimum 100 flight to that destination?

```{r, include=TRUE, message=TRUE, echo=FALSE}
q1 <- flight_data[origin == 'LGA', list(number_of_flights = .N, 
                                  mean_arrival_delay = mean(arr_delay, na.rm = T)), 
            by = dest][order(mean_arrival_delay)]

q1[number_of_flights > 100,][1]
```

## Q2. Which destination's flights were the most on time (avg arrival delay closest to zero) from LGA with minimum 100 flight to that destination?

```{r, include=TRUE, message=TRUE, echo=FALSE}
q2 <- q1[number_of_flights > 100, 
           list(dest, mean_arrival_delay, number_of_flights, 
                gicu = abs(mean_arrival_delay))][order(gicu)]

q2[, list(dest, number_of_flights, mean_arrival_delay)][1]
```

## Q3. Who is the manufacturer of the plane, which flights the most to CHS destination?

```{r, , include=TRUE, message=TRUE, echo=FALSE}
q3 <- flight_data[dest == 'CHS', list(number_flights = .N), 
                       by = list(tailnum, dest)][order(-number_flights)]
flight_data_planes <- data.table(nycflights13::planes)

merge(q3, flight_data_planes[, list(tailnum, manufacturer)])[order(-number_flights)][1]
```

## Q4. Which airline (carrier) flow the most by distance?

```{r, , include=TRUE, message=TRUE, echo=FALSE}
q4 <- data.table(nycflights13::airlines)

merge(flight_data[, list(total_distance_flown = sum(distance)), 
            by = carrier][order(-total_distance_flown)][1], q4)
```
## Q5. Plot the monthly number of flights with 20+ mins arrival delay!

```{r, message=TRUE, echo=FALSE, fig.width=8, fig.height=5, fig.align='center'}
q5 <- flight_data[arr_delay > 20, list(date = make_date(year, month))]

ggplot(q5, aes(x = date)) + geom_bar() +
  labs(y="Number of Flights", x = "Date")
```

## Q6. Plot the departure delay of flights going to IAH and the related day's wind speed on a scaterplot! Is there any association between the two variables? Try adding a linear model.

```{r,message=FALSE, echo=FALSE, fig.width=8, fig.height=5, fig.align='center', warning=FALSE}
flights_weather <- data.table(nycflights13::weather)

departure <- flight_data[dest == 'IAH', 
                       list(dep_delay, origin, dest, year, month, day)]
weather <- flights_weather[, list(mean_wind_speed = mean(wind_speed)), 
                               by = list(origin, year, month, day)]
q6 <- merge(departure, weather)

ggplot(q6, aes(dep_delay, mean_wind_speed)) + 
  geom_point() + 
  geom_smooth(method = 'lm') +
  labs(y="Mean Wind Speed", x = "Departure Delay") +
  theme_minimal()
```

## Q7. Plot the airports as per their geolocation on a world map, by mapping the number flights going to that destination to the size of the symbol!

```{r, message=FALSE, echo=FALSE, fig.width=8, fig.height=5, fig.align='center'}
map <- map_data('world')
ggplot() +
  geom_polygon(
    data = map, aes(x = long, y = lat, group = group), fill = 'gray', color = 'black') +
  geom_point(
    data = merge(
      airports, flight_data[, .(flights = .N), by = .(faa = dest)], by = 'faa'), aes(lon, lat, size = flights)) +
  theme_minimal() +
  theme(legend.position = 'top') +
  xlab('') + ylab('') 
```
